When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-07-09 code/Submissionv1.Rmd
commit changes

@@

My theory holds that when military credibility is high, a demand signal is created by the American public such that future interactions between military elites and the military institution are encouraged and welcomed. There are numerous ways in which these interactions might satisfy the public's demand to observe or hear from the military. Institutionally, the military might respond to the public's demand signal for increased engagement by participating in public ceremonies, parades, and other showcases of military prowess (i.e. at sporting venues). Individually, military elites, especially retired military officers, would be more eager to dialogue with the American public when military credibility is high. Press and media outlets would be more eager to showcase elite military voices when military credibility is high because they represent an institution that commands a high degree of public respect. When military credibility is low, the opposite will occur. Since demand for interactions between the American public and the military is muted, the institution will not participate in a high number of public showcases of military prowess because there will not be as strong of a demand signal sent by the public. In short, military credibility and respect for military expertise impacts the likelihood of the military institution interacting with the American public. When respect for the military is high, the military is likely to speak out and to interact with the public, and when respect for the military is low, the military is less likely to do so.

When political polarization is high, I contend that it ultimately places great strain on the concept and definition of loyalty because the distinction between what is considered morally right or wrong and partisan expression becomes blurred if not altogether removed. The result is a less unified military institution. I further contend that these mechanisms work as the logical and expected outworking of the characteristics of the polarized American political environment. While scholarly literature has looked at how the military faces pressure stemming from divided institutional loyalties [@betts_soldiers_1991], the dynamic of polarization promises to add even more pressure and scrutiny on the loyalty of individual officers and the military as a whole.

I contend that the the line between what is considered by individuals to be morally right and wrong and sentiments of partisan expression becomes blurred in highly polarized environments. As discussed above, political polarization has caused Americans to move further apart ideologically on an increasing number of issues [@layman_party_2006], including on core attitudes regarding issues such as race and immigration that are central to one's identity [@sides_identity_2018]. In this environment, individuals believe the stakes are raised to a very high level for a number pivotal concerns. Individuals in a polarized environment, I contend, also believe there are enormous consequences if the specific side with which they ally on any given issue loses to the other side because the other side is not just simply mistaken, but deeply and morally wrong.

This dynamic impacts military officers in a unique way because military officers, unlike regular citizens, operate under strict codes of loyalty all of the time. The American military officer, for example, takes an oath to support and defend the Constitution of the United States. In a polarized environment, two different officers may adopt completely different stances on a contentious issue, each believing the other is morally wrong and operating outside of the framework of the constitution. Each officer, then, might accuse the other of failing to be loyal to the ideals of the military. As a result, when polarization is high, the likelihood of partisan expression is high because in doing so, members of the military institution may feel that doing so is the right and loyal thing to do.

As a quick case in point, consider that in June 2020, following President Trump's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and politically striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@noauthor_letter_2020]. The letter's clear implication is that if you are serving in the Trump administration and if you graduated from the US Military Academy (i.e. Esper and Pompeo), you have forsaken your oath of office, the US Constitution, and the values that West Point stands for.

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table 1, below). The first is...the second is...The third is....and the fourth is....

@@

The second, and perhaps more important dynamic, is understanding the impacts of the polarized environment on the military's ability to perform its primary military task, i.e. fight and win the nation's wars. Scholars, military officers, and practitioners must better understand what impacts, if any, are placed on military and civilian leaders as a result of political polarization on the development and implementation of war goals, aims, and strategies, and how military leaders and civilian policymakers can overcome these hurdles to generate effective policies [@rapp_civil-military_2015].

>>>>

Figure 1, below, outlines the general theory and framework of the impact of political polarization and institutional credibility on military professionalism, and subsequently, on various military outcomes. The system of variables represented in Figure 1 constitute the "drive chain" that can lead to the words of retired military officers becoming militarized.

When military credibility is high, a demand signal is created by the American public such that future interactions between the military institution and the American public are encouraged and welcomed. From an institutional perspective, the military might respond to the public's demand signal for increased military interaction by increasingly participating in public ceremonies, parades, and other showcases of military prowess (at sporting venues, for instance). From an individual perspective, military elites, especially retired military officers, would be psychologically primed to increasingly dialogue with the American public when military credibility is high. By extension, other outlets such as the Press would also be primed to showcase military elites when military credibility runs high, because these elites represent an institution that commands a high degree of public respect.

When military credibility is low, I theorize the opposite will occur. Since demand for interactions between the American public and the military is muted when institutional credibility for the military is low, the institution will not engage with the public as frequently nor as intimately because the strong demand signal that existed when institutional credibility was high no longer exists. At the individual perspective, military elites will also not engage, nor will the Press ask these elites to, with the American public. In short, institutional credibility and respect for the military institution impacts the likelihood of the military institution and military elites interacting with the American public. Such interaction can take a variety of forms. When respect for the military is high, the military is likely to interact, speak to, and dialogue with the American public, and when institutional credibility is low, the military is less likely to do so.

@@

```
The degree of political polarization evident also has the potential to impact the professionalism of the military as an institution. First, because of the phenomenon known as "conflict extension" [@layman_party_2002], which refers to the fact that the political parties have grown further apart from each other on an increasing number of issues, political polarization raises the number and scope of issues to disagree about. Research has consistently shown that the polarizing politics present in the United States today has caused voters who identify as liberals to be more liberal across multiple issue categories, and that the same holds for conservatives. The result is that in the past, the prospect of cooperation on issues that were not polarized remained high, whereas today, the prospect for cooperation in general is lower [@layman_activists_2010]. As it concerns the military institution, "conflict extension" [@layman_party_2002] increases the likelihood that members of the military feel strongly about an increasing number of issues. In other words, I contend that today's polarized environment makes it more difficult for an institution that to be politically neutral when there are fewer issues to remain neutral over. At the individual level, the same is true: as the non-partisan space seems to shrink, individuals will be enticed to not remain neutral.

The second impact of political polarization on military professionalism deals with the fact that the issues which are becoming polarized are not merely matters of simple policy preferences, but substantive issues which are directly related to notions of a person's core identity, such as race and immigration [@sides_identity_2018], religion [@white_values_2002], and the future of the country [@mann_its_2012]. These issues do not lend themselves to compromise. They are, for many people, using an economics phrase, "indivisible goods." This also explains the notion of "negative partisanship," which refers to the very strong negative feelings that exist on each side of the political spectrum for those on the other [@abramowitz_great_2018, 5]. Citizens feel and believe that the stakes are raised to an extremely high level for a number of pivotal issue areas, and that there are enormous consequences if the specific side with which they ally on any given issue loses to the other side. Polarization, then, raises the salience of right and wrong in a moral sense, which increases the odds of expressing partisan sentiment.

The third and most important way in which political polarization impacts military professionalism, I contend, extends from the first two reasons outlined above. First, because there are less issue areas in general to remain politically neutral about, and second, because issues are increasingly viewed in terms of moral right and wrong, political polarization impacts and distorts the notions of two vital character traits: loyalty and duty. As any soldier knows, few traits are more sacred in the military than loyalty and duty. In a very real sense, the military professional and the institution at large see loyalty and duty as incentives. By this, I simply mean that military professionals pride themselves on living up to mantras and phrases which extol the virtues of "doing the right thing," "leading with integrity," and, as the West Point Cadet Prayer so elegantly states, choosing the "harder right over the easier wrong."

Political polarization obfuscates these ideals, however, and places greater responsibility for interpreting the meaning and the associated action of these ideals on the individual, and less on the military institution. It is in this way, I contend, that political polarization raises the chance of the military professional acting in such a way that simultaneously expresses partisan sentiment, because in doing so, the military professional believes he or she is doing what is morally right.

This is why in a politically polarized environment, two different military officers may adopt completely different opinions of a contentious issue, each believing that the other's opinion is wrong. Should the officer act on that opinion, the other might believe he or she is operating outside of established military frameworks, the constitution, or any number of structures. Each officer, then, might accuse the other of failing to be loyal to the ideals of the military. As a result, when polarization is high, the likelihood of partisan expression is high because in doing so, members of the military institution may feel that doing so is the right and loyal thing to do.

As a quick case in point, consider that in June 2020, following President Trump's largely unpopular move of displacing protesters from Lafayette Square, a letter published by West Point alumni expressly called out prominent West Point alumni serving in the Trump administration. Though it did not mention individuals by name, the letter was intended and acknowledged by media as directed towards the Secretary of Defense, Mark Esper, and the Secretary of State, Mike Pompeo, who are both graduates of West Point's Class of 1986 [@gaouette_group_nodate]. The charges of the letter were both damning and politically striking: "...We are concerned that fellow graduates serving in senior-level, public positions are failing to uphold their oath of office and their commitment to Duty, Honor, Country"[@noauthor_letter_2020]. The letter's clear implication is that if you are serving in the Trump administration and if you graduated from the US Military Academy (i.e. Secretaries Esper and Pompeo), you have forsaken your oath of office, the US Constitution, and the values that West Point stands for.

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table 1, below). The first is when political polarization and military credibility are both low. When these conditions hold, there is no anticipated drop in military professionalism by the military institution nor its individual elites. With these conditions, the military does not have an out-sized platform to interact with the American public, nor are there outstanding incentives for the military to raise significant disagreement over any domestic political issues.

The second possible set of conditions is when political polarization is low and institutional credibility is high. When these set of conditions exist, the military has a platform to interact with the public, as the public highly respects the military institution. The military enjoys this high level of trust and does not want to lose it. Further, because the degree of political polarization is low, there is little incentive for the military as an institution or military elites to use the trust they enjoy with the American public to adopt or denounce an issue associated with any partisan entity. The high institutional credibility increases the chance of the military interacting with the public in an unprofessional way, but the low political polarization removes the incentives of doing so.

The third type occurs when political polarization is high, but institutional credibility of the military is low. Under these conditions, the military does not command the respect of the American public like it does when credibility of the institution is high, and thus, does not possess a considerable outlet to interact with the American public. However, the incentives of loyalty and duty still loom large in the minds of some military elites as a result of the high degree of political polarization, and thus, a dip in professionalism is still possible.

The final set of conditions exists when political polarization and institutional credibility are high. It is in this world that I contend the American military finds itself in now in 2020. Under these conditions, the military faces dual dangers. High institutional credibility risks the military institution unilaterally acting or another body co-opting the military institution to speak out an a number of issue areas, and simultaneously, the highly polarized environment risks the military advocating for or denouncing partisan sentiments in the name of upholding the very traits it is most adhered to - loyalty and duty. Indeed, this set of conditions comprise a very dangerous world for the military institution and the military professional. I

@@

The second, and perhaps more important dynamic, is understanding the impacts of the polarized environment on the military's ability to perform its primary military task, i.e. fighting and winning the nation's wars. Scholars, military officers, and practitioners must better understand what impacts, if any, are placed on military and civilian leaders as a result of political polarization on the development and implementation of war goals, aims, and strategies, and how military leaders and civilian policymakers can overcome these hurdles to generate effective policies [@rapp_civil-military_2015]. While scholarly literature has looked at how the military faces pressure stemming from divided institutional loyalties [@betts_soldiers_1991], none have explored the unique ways in which polarization as a domestic political force shapes individual and institutional behavior of the military.
